---
title: "STAT112 Final Project"
output: html_document
author: "Hayley Hadges"
---

# INTRODUCTION:

### RESEARCH QUESTIONS:

1. How does the cost of college differ for income level?

2. How has the cost of college changed over time?

3. Is the cost of college worth the payout?

4. Does it really matter where you receive your degree?


# DATA: 
I found these datasets on Tidy Tuesday, 2020-03-10, called "College tuition, diversity, and pay". The data comes from many different sources but originally came from the US Department of Education, and the massive amount of data that was filtered down to a few tables.

# ANALYSIS:

```{r}
library(tidyverse)
library(geofacet)
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes) 
library(openintro) 
library(ggridges)
```

```{r}
tuition_cost <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/tuition_cost.csv')

tuition_income <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/tuition_income.csv') 

salary_potential <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/salary_potential.csv')

historical_tuition <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/historical_tuition.csv')

diversity_school <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/diversity_school.csv')
```

```{r}
ht <-
historical_tuition %>% 
  filter(tuition_type == "4 Year Constant",
         type != "All Institutions") %>% 
  separate(year,c("start_year", "end"), "-" )

ht %>% 
  ggplot(aes(x = start_year, y = tuition_cost, group = type, color = type)) +
  geom_line()
```

```{r}
avg_state_tuition <-
tuition_cost %>% 
  filter(degree_length == "4 Year",
         type != "For Profit") %>% 
  group_by(state, type) %>% 
  summarise(avg_tuition = mean(in_state_tuition))
```

```{r, fig.height= 12, fig.width=16}
income <-
tuition_income %>% 
  mutate(state_name = state.name[match(state, state.abb)]) %>% 
  group_by(state_name, year, income_lvl) %>% 
  summarise(percent_total_tuition = mean(net_cost)/mean(total_price),
            avg_net_price = mean(net_cost))

income2 <-
tuition_income %>% 
  group_by(state, year, income_lvl) %>% 
  summarise(percent_total_tuition = mean(net_cost)/mean(total_price),
            avg_net_price = mean(net_cost))
```

```{r}
income2 %>% 
  ggplot(aes(x = year, y = avg_net_price, color = income_lvl)) +
  geom_line() + 
  facet_geo(~ state)

income2 %>% 
  ggplot(aes(x = year, y = percent_total_tuition, color = income_lvl)) +
  geom_line() + 
  facet_geo(~ state)


income2 %>% 
  filter(state == "MN") %>% 
  ggplot(aes(x = year, y = avg_net_price, color = income_lvl)) +
  geom_line()

income2 %>% 
  filter(state == "MN") %>% 
  ggplot(aes(x = year, y = percent_total_tuition, color = income_lvl)) +
  geom_line()
```


```{r}
percent_div <-
diversity_school %>% 
  mutate(percent_minority = (enrollment / total_enrollment) * 100)
```


```{r}
ht2010 <-
ht %>% 
  filter(start_year >= 2010)
```

```{r}
avg_potential <-
salary_potential %>% 
  group_by(state_name) %>%
  summarise(avg_early_pay = mean(early_career_pay),
            avg_mid_pay = mean(mid_career_pay),
            avg_better_world = mean(make_world_better_percent, na.rm=TRUE),
            avg_stem = mean(stem_percent))
```

```{r}
injwdc <-
tuition_cost %>%
  left_join(tuition_income %>%
  filter(year == 2018,
         campus == "On Campus"), 
         by = "name")
```


```{r}
new_data <-
salary_potential %>% 
  left_join(tuition_income, by = c("name" = "name")) %>% 
  filter(year == 2018,
         campus == "On Campus")

new_data2 <-
salary_potential %>% 
  left_join(income, by = c("state_name" = "state_name")) %>% 
  filter(year == 2018)
```

```{r}
new_data2 %>% 
  ggplot() +
  geom_jitter(aes(x = avg_net_price, y = mid_career_pay, color = income_lvl)) 

new_data %>% 
  group_by(name) %>% 
   ggplot() +
  geom_jitter(aes(x = total_price, y = mid_career_pay)) 

```

```{r, fig.height=15, fig.width= 18}
salary_potential %>% 
  ggplot(aes(y=early_career_pay, x=fct_reorder(state_name, early_career_pay, mean))) + 
    geom_boxplot(position="dodge", alpha=0.5, outlier.colour="transparent") +
  coord_flip()

salary_potential %>% 
  filter(state_name == "Minnesota" |
           state_name == "California" |
           state_name == "Washington") %>% 
  ggplot(aes(y=mid_career_pay, x=fct_reorder(state_name, mid_career_pay, mean), fill = state_name)) + 
    geom_boxplot(position="dodge", alpha=0.5, outlier.colour="transparent") +
  coord_flip()
```

```{r}
states_map <- map_data("state")

avg_potential %>% 
  mutate(state = str_to_lower(state_name)) %>% 
 # str_replace_all(state, "-", "") %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state,
               fill = avg_early_pay)) +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  scale_fill_gradient2() +
  theme_map()


avg_potential %>% 
  mutate(state = str_to_lower(state_name)) %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state,
               fill = avg_mid_pay)) +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  scale_fill_gradient2() +
  theme_map()
```

```{r}
tuition_income %>% 
  filter(year == 2018,
         state == "MN" |
           state == "CA" |
           state == "WA") %>% 
  group_by(name, state) %>% 
  ggplot(aes(x = net_cost, y = income_lvl, fill = income_lvl)) +
    geom_density_ridges() +
  theme_ridges() +
  facet_wrap(~state, ncol = 1)


tuition_income %>% 
  filter(year == 2018,
         state == "MN" |
           state == "CA" |
           state == "WA",
         income_lvl == "0 to 30,000") %>% 
  group_by(name, state) %>% 
  ggplot(aes(x = net_cost, y = state, fill = state)) +
    geom_density_ridges() +
  theme_ridges() 

tuition_income %>% 
  filter(year == 2018,
         state == "MN" |
           state == "CA" |
           state == "WA",
         income_lvl == "0 to 30,000") %>% 
  group_by(name, state) %>% 
ggplot(aes(y=net_cost, x=state)) + 
    geom_violin(position="dodge") +
  coord_flip()
```

```{r}
tuition_cost %>% 
  filter(degree_length == "4 Year",
         type == "Private") %>% 
   mutate(tuition_dif = out_of_state_tuition - in_state_tuition) %>% 
  group_by(state)
         
tuition_cost %>% 
  filter(degree_length == "4 Year",
         type == "Public",
         state == "Minnesota" |
           state == "California" |
           state == "Washington") %>% 
  mutate(tuition_dif = out_of_state_tuition - in_state_tuition) %>% 
  group_by(state, type) %>% 
 ggplot(aes(x = tuition_dif, y = fct_reorder(state, tuition_dif, mean))) +
    geom_density_ridges() +
  theme_ridges() 
```

```{r}
tuition_cost %>% 
  filter(degree_length == "4 Year",
         type == "Public",
         state == "Minnesota" |
           state == "California" |
           state == "Washington") %>% 
  mutate(tuition_dif = out_of_state_tuition - in_state_tuition) %>% 
  group_by(state, type) %>%
ggplot(aes(y=tuition_dif, x=fct_reorder(state, tuition_dif, mean))) + 
    geom_violin(position="dodge") +
  coord_flip()
```


```{r}
diversity <-
diversity_school %>% 
  filter(state == "Minnesota" |
           state == "California" |
           state == "Washington") %>% 
  pivot_wider(id_cols = name:state,
              names_from = category,
              values_from = enrollment) %>% 
  group_by(state) %>% 
  mutate(avg_total_enrollment = mean(total_enrollment),
         avg_total_min = mean(`Total Minority`),
         avg_native = mean(`American Indian / Alaska Native`),
         avg_asian = mean(`Asian`),
         avg_black = mean(`Black`),
         avg_hispanic = mean(`Hispanic`),
         avg_islander = mean(`Native Hawaiian / Pacific Islander`),
         avg_multiple = mean(`Two Or More Races`),
         avg_white = mean(`White`)) 

diversity %>% 
  ggplot(aes(x = state, y = total_enrollment, fill = )) +
  geom_bar(stat = "identity")
  

```

```{r}
diversity2 <-
diversity_school %>% 
  filter(state == "Minnesota" |
           state == "California" |
           state == "Washington") %>% 
  filter(!category %in% c("Women", "Unknown", "Non-Resident Foreign", "Total Minority"))

diversity2 %>%  
   ggplot(aes(x = state, y = enrollment, fill = category)) +
  geom_bar(position = "fill", stat = "identity")
```


```{r}
library(states)
state_vector <- 
  salary_potential %>% 
  distinct(state_name)

#state_list <- list(state_vector)

names(state_vector) <-state.name


```